### 2017-07-14 Blog/07-WhatIsInformationEntropy.tex
Blog post review

The concept of entropy in physics is unfortunately associated with a certain amount of vagueness. While mathematically is well defined, I have never heard a crisp conceptual understanding (e.g. it measures "disorder", it has to do with how much the system is homogeneous, ...). Entropy in information theory, or Shannon's entropy, instead has a crisp intuitive meaning and therefore I cringe when I hear some defining it as "lack of information" or any other vague description. So, let's see if we can do better.

@@

Suppose you have a bag with 40 balls. 20 are green and 20 are red. Suppose you pick a ball and I want to know the color. Suppose I can only ask you yes or no questions. How many questions do I have to ask you? One suffices: "is it red?". If the answer is yes, I know it's red, if the answer is no, I know it's green. We can rephrase by asking: how many one-zero choices? That is: how many bits of information you need to give me? The answer is one bit: 0 for green and 1 for red. That is: all the information can be encoded into one zero-one choice.

Let's do it again in another case. Suppose all 40 balls are green. How many bits of information? The answer is zero bits: I already know you are going to get a green ball.

Suppose there are 10 balls for each of these color: green, red, yellow and blue. How many bits of information? We can set 00 for green, 01 for red, 10 for yellow and 11 for blue. That is two zero-one choices: it's two bits of information.

@@

First of all, information entropy has nothing to do with knowledge (e.g. what we know about the system or what can be known). This misunderstanding arises because the concept can be applied to probability distribution that represent our knowledge of the system, and in some cases it may relate to knowledge. For example, if there are four boxes and I do not know in which one the ball is hiding, I have a uniform distribution on four cases which has 2 bits of information entropy. Conversely, if I know where it is, the entropy is 0 bits. So, in this case more entropy correspond to less knowledge. But this idea does no generalize.

@@

This hopefully helps you understand what information entropy measures: how many bits of informations are required on average to identify an element in a distribution. Not "knowledge" or "lack of information" or "disorder" or any other vague concept. In the end, information entropy is a number: to understand it you need to understand what that number quantifies and in what units.

>>>>

The concept of entropy in physics is unfortunately associated with a certain amount of vagueness. While mathematically is well defined, I have never heard a crisp conceptual understanding (e.g. it measures "disorder", it has to do with how much the system is homogeneous, ...). Entropy in information theory, or Shannon's entropy, instead has a crisp intuitive meaning and therefore I cringe when I hear people describing it as "lack of information" or any other vague description. So, let's see if we can do better.

@@

Suppose you have a bag with forty balls. Twenty are green and twenty are red. Suppose you pick a ball and I want to know the color. Suppose I can only ask you yes or no questions. How many questions do I have to ask you? One suffices: "is it red?". If the answer is yes, I know it's red, if the answer is no, I know it's green. We can rephrase by asking: how many one-zero choices? That is: how many bits of information you need to give me? The answer is one bit: 0 for green and 1 for red. That is: all the information can be encoded into a single zero-one choice.

Let's do it again in another case. Suppose all forty balls are green. How many bits of information? The answer is zero bits: I already know you are going to get a green ball.

Suppose there are ten balls for each of these color: green, red, yellow and blue. How many bits of information? We can set 00 for green, 01 for red, 10 for yellow and 11 for blue. Four cases can be encoded into two zero-one choices: it's two bits of information.

@@

First of all, information entropy has nothing to do with knowledge (e.g. what we know about the system or what can be known). This misunderstanding arises because the concept can be applied to probability distribution that represent our knowledge of the system, and in some cases it may relate to knowledge. For example, if there are four boxes and I do not know in which one the ball is hiding, I have a uniform distribution on four cases which has 2 bits of information entropy. Conversely, if I know where it is, the entropy is 0 bits. So, in this case more entropy correspond to less knowledge. But this idea does not generalize.

@@

At the end of the day, information entropy is a number: to understand it you need to understand what that number quantifies and in what units. If do not have a clear idea what that number is measuring, you do not understand what it is. That number measures how many bits of informations are required on average to identify an element in a distribution. Not "knowledge" or "lack of information" or "disorder" or any other vague concept.
